---
title: "RNA Structure changes due to editing"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlights: pygments
    self_contained: true
    number_sections: yes
    theme: readable
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: yes
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  echo=T, # shows code
  cache=F,message = F, warning = F
)
set.seed(31)
```

```{r, Functions, echo=T}
rm(list=ls())
source("A:/work/scripts/SchumacherLab/vars.R")
source("A:/work/scripts/SchumacherLab/functions.R")

```


# Structure comparison of the RNA-hairpins 
- Secondary structure predicted using RNAfold software (ref-1)
- Color code on the figure can be read as high base pairing likelyhood
- clustered events across genic regions are considered in the analyses (with overall locus length <1kb)
- Please read the figure heading in following manner
    * Gene 
    * Annotation of the clustered editing locus
    * numbner of edits in the cluster;
- Note:The script computes a well-de???nedness measure called "positional entropy" and encodes it as color hue
- Red (low entropy, well-de???ned) via green to blue and violet (high entropy, ill-de???ned).
- Refer for more details (https://www.tbi.univie.ac.at/RNA/tutorial/)

# spreadsheet-report

```{r, p1, echo=T,fig.width=8,fig.height=6}
  myf<- function(path="A:/work/Kreidberg lab/Valerie/PAPER/PSVisulaization/Ref/DRE_Clustered_RefSeqs_Genomic_OUT.fa"){
    ref <- read.delim(path,header=F,stringsAsFactors = F)
    rh <-  data.frame(id=ref[seq(1,nrow(ref),6),])
    rh$energy <- ref[seq(3,nrow(ref),6),]
    rh$energy <- gsub(".*?\\(","",rh$energy)
    rh$energy <- gsub("\\)","",rh$energy)
    rh$energy <- as.numeric(rh$energy)
    rh$id <- gsub(">","",rh$id)
    rm(ref)
    rh
  }
  
  ref <- myf(path="A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Cntr/33hc_DRE_Clustered_RefSeqs_Genomic_OUT.fa")
  alt <- myf(path="A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Exp/33hc_DRE_Clustered_AltSeqs_Genomic_OUT.fa")
  names(ref)[2] <- "ref_energy"
  names(alt)[2] <- "alt_energy"
  rh <- merge(ref,alt,by="id")
  rh$pic <- paste0(gsub("\\|","_",rh$id),"_rel.jpg")
  rh$gene <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$ncluster <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$FDR <- gsub("\\|\\S*","",rh$id)
  rh$FDR <- formatC(rh$FDR, format = "e", digits = 2)  
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$Anno <- gsub("\\|\\S*","",rh$id)
  rh$Anno <- gsub("_prime_","'",rh$Anno)
  rh$Anno <- gsub("_region","",rh$Anno)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$medainChange <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$comparison <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$id <- NULL
  rh <- rh[,c("gene", "ncluster", "FDR", "Anno", "medainChange", "comparison","ref_energy", "alt_energy", "pic")]

  rh$score <- round((rh$ref_energy- rh$alt_energy)*100/abs(rh$ref_energy),2)
  rh <- rh[order(rh$score,decreasing = T),]
  
  rh1 <- rh[,c("gene", "ncluster", "FDR", "Anno", "medainChange", "comparison", 
"ref_energy", "alt_energy", "score")]
  names(rh1) <- c("gene", "EditsInClusters", "FDR_EditLevels", "Anno", "medainEditChange", "comparison", 
                "Control_energy", "D7/D14_energy", "score")
  
  load(paste0(DATADIR,"33hc_Spreadsheet_HCEditsWithCompleteAnno.RData"))
  SPN <- SPN[,c("gene","Control_log2Expn","D7_log2Expn","D14_log2Expn","GOAnno","link")] %>% unique
  rh1 <- merge(rh1, SPN,by="gene",all.x=T)

  datatable(rh1, extensions = 'Buttons',options = list(searching = TRUE,pageLength = 10,dom = 'Bfrtip', buttons = c('excel', "csv")))

```

# Figures
- The structure of edited mRNA (Pank3) is presented for control, D7 and D14 condition, on top
- While, the paired boxplot at the bottom summarizes changes in free-energy of all structures.
- Out of 84 (48 for D7 and 36 for D14) instances, I selected those where the editing is elevated in D7 or D14 compared to Control samples. 
- In the boxplots, I plotted changes in energy for premRNA structures with respect to their structures observed in control samples. 
- The structures in control are therefore based out at 0 and the changes are quantified against the baseline (red dotted line)
- As you can see, overall changes suggest stabilization of the RNA structures (boxes moving below 0-line in D7 and D14). However, this change is not statistically significant! (reported by ns on the top) 


```{r, p2, echo=T,fig.width=8,fig.height=8}
  myf<- function(path="A:/work/Kreidberg lab/Valerie/PAPER/PSVisulaization/Ref/DRE_Clustered_RefSeqs_Genomic_OUT.fa"){
    ref <- read.delim(path,header=F,stringsAsFactors = F)
    rh <-  data.frame(id=ref[seq(1,nrow(ref),6),])
    rh$energy <- ref[seq(3,nrow(ref),6),]
    rh$energy <- gsub(".*?\\(","",rh$energy)
    rh$energy <- gsub("\\)","",rh$energy)
    rh$energy <- as.numeric(rh$energy)
    rh$id <- gsub(">","",rh$id)
    rm(ref)
    rh
  }
  
  ref <- myf(path="A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Cntr/33hc_DRE_Clustered_RefSeqs_Genomic_OUT.fa")
  alt <- myf(path="A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Exp/33hc_DRE_Clustered_AltSeqs_Genomic_OUT.fa")
  names(ref)[2] <- "ref_energy"
  names(alt)[2] <- "alt_energy"
  rh <- merge(ref,alt,by="id")
  rh$pic <- paste0(gsub("\\|","_",rh$id),"_rel.jpg")
  rh$gene <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$ncluster <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$FDR <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$Anno <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$medainChange <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$comparison <- gsub("\\|\\S*","",rh$id)
  rh$id <- gsub("^\\S*?\\|","",rh$id)
  rh$id <- NULL
  rh <- rh[,c("gene", "ncluster", "FDR", "Anno", "medainChange", "comparison","ref_energy", "alt_energy", "pic")]
  rhx <- rh
  
  load(paste0(DATADIR,"33hc_DRE_DE_matrix.RData"))
  r1 <- res[res$comparison=="D7_control" & res$clustered>0 & !is.na(res$tracking_id) & res$adj.P.Val<0.05,]
  r1 <- r1[r1$D7>r1$control,]
  rh1 <- rh[rh$comparison=="D7_control" & rh$gene%in%r1$gene,]
  r1 <- res[res$comparison=="D14_control" & res$clustered>0 & !is.na(res$tracking_id) & res$adj.P.Val<0.05,]
  r1 <- r1[r1$D14>r1$control,]
  rh2 <- rh[rh$comparison=="D14_control" & rh$gene%in%r1$gene,]
  rh <- rbind(rh1,rh2)
  rm(rh1,rh2,r1)  
  
  rh <- rh[rh$medainChange>0,c("gene","comparison","ref_energy","alt_energy")]
  #rh <- rh[rh$ref_energy>rh$alt_energy,]
  rh$comparison <-gsub("_control","",rh$comparison)
  rh <- reshape2::dcast(rh,gene+ref_energy ~ comparison, value.var = "alt_energy")
  names(rh)[2:4] <- c("Control","Day-14","Day-7")
  
  rh1 <- rh[,c(1,2,4)]
  rh1 <- rh1[complete.cases(rh1),]
  rh1[,3] <- ((rh1[,3]-rh1[,2])*100)/abs(rh1[,2])
  rh1[,2] <- 0
  rh1 <- melt(rh1,measure.vars = names(rh1)[2:3])
  p4 <- ggpaired(rh1, x = "variable", y = "value",color = "variable", 
                 add = "jitter",line.color = "gray", line.size = 0.4)+gg_aes+
    scale_color_manual(values=cols)+scale_fill_manual(values=cols)+
    geom_hline(yintercept = 0,col="red",linetype="dotted",size=1.2)+
    stat_compare_means(label = "p.signif", method = "t.test",paired = T,ref.group = "Control")+
    xlab("")+ylab("% change in energy\n[ (Control-D7)/Control ]")
  
  rh1 <- rh[,c(1,2,3)]
  rh1 <- rh1[complete.cases(rh1),]
  rh1[,3] <- ((rh1[,3]-rh1[,2])*100)/abs(rh1[,2])
  rh1[,2] <- 0
  rh1 <- melt(rh1,measure.vars = names(rh1)[2:3])
  p5 <- ggpaired(rh1, x = "variable", y = "value",color = "variable", 
            add = "jitter",line.color = "gray", line.size = 0.4)+gg_aes+
    scale_color_manual(values=cols)+scale_fill_manual(values=cols)+
    geom_hline(yintercept = 0,col="red",linetype="dotted",size=1.2)+
    stat_compare_means(label = "p.signif", method = "t.test",paired = T,ref.group = "Control")+
    xlab("")+ylab("% change in energy\n[ (Control-D14)/Control ]")
  
  p1 <- ggdraw() + draw_image(paste0("A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Cntr/",rhx$pic[50]),scale = 1)+
    ggtitle(paste0("Control\n",rhx$gene[50],"\n(",rhx$ref_energy[50],"  kcal/mol)" ) )+
    theme(plot.title = element_text(hjust = 0.5))
  p2 <- ggdraw() + draw_image(paste0("A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Exp/",rhx$pic[50]),scale = 1)+  
    ggtitle(paste0("Day-7\n",rhx$gene[50],"\n(",rhx$alt_energy[50],"  kcal/mol)" ) )+
    theme(plot.title = element_text(hjust = 0.5))
  p3 <- ggdraw() + draw_image(paste0("A:/work/Kreidberg lab/Valerie/PAPER/33hc_PSVisulaization/Cntr/",rhx$pic[50]),scale = 1)+  
    ggtitle(paste0("Day-14\n",rhx$gene[50],"\n(",rhx$ref_energy[50],"  kcal/mol)" ) )+
    theme(plot.title = element_text(hjust = 0.5))
  
  lay <- rbind(c(1,1,2,2,3,3),
               c(4,4,4,5,5,5))
  grid.arrange(p1,p2,p3,p4,p5,heights=c(5,4), layout_matrix = lay)
  

```


# References
1. Mathews DH, Disney MD, Childs JL, Schroeder SJ, Zuker M, Turner DH. (2004) Incorporating chemical modification constraints into a dynamic programming algorithm for prediction of RNA secondary structure. Proc Natl Acad Sci U S A 101(19):7287-92.



# Sessioninfo
```{r, sessioninfo, echo=T}
sessionInfo()
```















